Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly

### 2014-10-20 TACO-paper-v2/coffee.tex
Minor edits, including the splitting stuff

Splitting an expression (henceforth \emph{split}) has, however,
several drawbacks. Firstly, it increases the number of accesses to $A$
in proportion to the ``split factor,'' which is the number of
sub-expressions produced. Also, depending on how splitting is done, it
can lead to redundant computation. For example, the number of times
the product $det*W3[i]$ is performed is proportional to the number of
sub-expressions, as shown in the code snippet. Further, it increases
loop overhead, for example through additional branch
instructions. Finally, it might affect register locality: for
instance, the same array could be accessed in different
sub-expressions, requiring a proportional number of loads be
performed. This is not the case for the Helmholtz
example. Nevertheless, as shown in Section~\ref{sec:perf-results}, the
performance gain from improved register reuse along inner dimensions
can still be greater, especially if the split factor and the splitting
itself use heuristics to minimize the aforementioned issues.

@@

In this paper, we have presented the study and systematic performance
evaluation of a class of composable cross-loop optimizations for improving
arithmetic intensity in finite element local assembly kernels, and their
integration in a novel compiler, COFFEE. In the context of automated code
generation for finite element local assembly, COFFEE is the first compiler
capable of introducing low-level optimizations to maximize instruction-level
parallelism, register locality and SIMD vectorization. Assembly kernels have
particular characteristics. Their iteration space is usually very small, with
the size depending on aspects like the degree of accuracy one wants to
reach (polynomial order of the method) and the mesh discretization employed.
The data space, in terms of number of arrays and scalars required to evaluate
the element matrix, grows proportionately with the complexity of the
finite element problem. COFFEE has been developed taking into account
all of these degrees of freedom, based on the idea that reducing the
problem of local assembly optimization to a fixed sequence of
transformations is far too superficial if close-to-peak performance
needs to be reached. The various optimizations overcome limitations of
current vendor and research compilers. The exploitation of domain
knowledge allows some of them to be particularly effective, as
demonstrated by our experiments on two state-of-the-art Intel
platforms. Further work includes a comprehensive study about
feasibility and constraints on transforming kernels into a sequence of
calls to external linear algebra libraries. COFFEE supports all of the
problems expressible in Firedrake, and is already integrated with
this framework. The generality and the applicability of the proposed
code transformations to other domains has also been discussed.

>>>>

Splitting an expression (henceforth \emph{split}) has, however,
several drawbacks. Firstly, it increases the number of accesses to $A$
in proportion to the ``split factor,'' which is the number of
sub-expressions produced. Also, depending on how splitting is done, it
can lead to redundant computation. For example, the number of times
the product $det*W3[i]$ is performed is proportional to the number of
sub-expressions, as shown in the code snippet. Further, it increases
loop overhead, for example through additional branch
instructions. Finally, it might affect register locality: for
instance, the same array could be accessed in different
sub-expressions, requiring a proportional number of loads be
performed; this is not the case for the Helmholtz
example, though. Nevertheless the performance gain from improved register reuse 
can still be greater if suitable heuristics are used. Our approach 
consists of traversing the expression tree and recursively splitting it into 
multiple sub-expressions if the number of innermost-loop independent 
variables is larger than a certain threshold. This is elaborated in the next section, 
and validated against empirical search in Section~\ref{sec:perf-results-split}.

% Nevertheless, as shown in Section~\ref{sec:perf-results}, the
%performance gain from improved register reuse along inner dimensions
%can still be greater, especially if the split factor and the splitting
%itself use heuristics to minimize the aforementioned issues.

@@

In this paper, we have presented the study and systematic performance
evaluation of a class of composable cross-loop optimizations for improving
arithmetic intensity in finite element local assembly kernels, and their
integration in a novel compiler, COFFEE. In the context of automated code
generation for finite element local assembly, COFFEE is the first compiler
capable of introducing low-level optimizations to maximize instruction-level
parallelism, register locality and SIMD vectorization. Assembly kernels have
particular characteristics. Their iteration space is usually very small, with
the size depending on aspects like the degree of accuracy one wants to
reach (polynomial order of the method) and the mesh discretization employed.
The data space, in terms of number of arrays and scalars required to evaluate
the element matrix, grows proportionately with the complexity of the
finite element problem. COFFEE has been developed taking into account
all of these degrees of freedom, based on the idea that reducing the
problem of local assembly optimization to a fixed sequence of
transformations is far too superficial if close-to-peak performance
needs to be reached. The various optimizations overcome limitations of
current vendor and research compilers. The exploitation of domain
knowledge allows some of them to be particularly effective, as
demonstrated by our experiments on two state-of-the-art Intel
platforms. Further work includes a comprehensive study about
feasibility and constraints on transforming kernels into a sequence of
calls to external linear algebra libraries. As suggested by the performance evaluation, an analysis of more sophisticated approaches for expression splitting also looks promising. COFFEE supports all of the
problems expressible in Firedrake, and is already integrated with
this framework. The generality and the applicability of the proposed
code transformations to other domains has also been discussed.
